Indirect elicitation as a necessary condition for consistent surrogate losses

### 2020-06-02 neurips-20/cvx-flats.tex
pass over 6

%commented out Jun 1 2020
%Theorem~\ref{thm:consistent-implies-indir-elic} allows us to use convex elicitation complexity as a tool to understand efficiency of consistent convex surrogates for a given property, which is more often what is given in a continuous estimation setting.
%For example, when one wants to learn an $\alpha$-quantile, we start with the property rather than a loss.
%In the literature, pinball loss $L(r,y) = (r-y)(\ones_{r \geq y} - \alpha)$ typically appears without explanation or justification as to why it is relevant in relation to the $\alpha$-quantile.
%Elicitation teaches us why pinball loss is consistent for learning a quantile as the pinball loss elicits the $\alpha$-quantile.
Shifting away from discrete predictions, we turn our attention to continuous estimation settings.
More often, the continuous estimation setting does not start with a target loss, but instead with a statistic of the data that we want to estimate, such as the conditional mean, variance, or entropy.
We use Theorem~\ref{thm:cvx-flats} in conjunction with results from~\citet{frongillo2018elicitation} to prove Theorem~\ref{thm:bayes-risk-lower-bound}, which yields new bounds on the convex elicitation complexity of properties like the variance and entropy.
\botodo{The pinball stuff felt a bit too much of a digression, I think. Jessie, could you rewrite this paragraph to say what readers should expect in the upcoming section? Perhaps mention quadrants explicitly to help readers follow the flow from the previous section. We can save the pinball text for a better opportunity.\jessie{Attempting... old paragraph commented out.}}

Theorem~\ref{thm:cvx-flats} also addresses one major open question from~\citet{frongillo2015elicitation,frongillo2018elicitation}\jessiet{which citation?}: \emph{what are lower bounds on convex elicitation complexity?}
While Theorem~\ref{thm:cvx-flats} is a statement about consistent surrogates, the heart of it is actually about indirect property elicitation.
Thus, the result also applies to extend previous results~\citep{frongillo2015elicitation,frongillo2018elicitation} to provide new bounds on elicitation complexity for certain classes of properties in Theorem~\ref{thm:bayes-risk-lower-bound}.

@@

  By Theorem~\ref{thm:cvx-flats}, as $\hat L$ elicits $\hat\Gamma$ and indirectly elicits $\Gamma$, there is a level set $\hat\Gamma_{\hat r}$ and flat $\hat F\subseteq\affhull(\simplex)$ containing $p$ such that $\codim(\hat F) \leq k$ and $S := \hat F \cap \simplex \subseteq \hat\Gamma_{\hat r} \subseteq \Gamma_r$.
  \raft{This last statement does follow from the proof of our main theorem, but it takes a second to realize that we show the chain of inclusions. \jessie{Tried to clarify in that proof}}
  Let $F = \affhull(\Gamma_r)$ and recall $\codim(F)=d$.
  % Let $F' = \hat F \cap \affhull(\simplex)$, and observe the following: $p\in F'\cap\relint\simplex$, $F' \subseteq \affhull(\simplex)$, $S=F'\cap\simplex\subseteq\Gamma_r$, and $\codim(F') \leq k+1$.
  % the next \hat Fs used to be F's
  As $p\in \hat F\cap\relint\simplex$, Lemma~\ref{lem:affhull-relint} gives $\affhull(S) = \affhull(\hat F\cap \simplex) = \hat F$.
  Now as $S \subseteq \Gamma_r$, we also have $\hat F = \affhull(S) \subseteq \affhull(\Gamma_r) = F$,
  implying $\codim(\hat F) \geq \codim(F)$.

@@

\subsection{Examples}

As a warm-up, let us see how to show $\eliccvx(\Var)=2$ whenever $|\Y|\geq 3$, meaning the lowest dimension of a convex loss to estimate conditional variance is 2.
It is interesting to note that, to the best of our knowledge, even this simple bound is novel.
While intuitively obvious, the lower bound of 2 is not trivial.
In particular, the well-known fact that the variance is not elicitable does not yield this lower bound, as it does not rule out the variance being a link of a real-valued convex-elicitable property; cf.~\citet{frongillo2018elicitation}.

Let $\Y\subseteq\reals$ be a finite set with $n=|\Y|$.
Observe that the variance of $Y$ is the Bayes risk of squared loss $L(r,y) = (r-y)^2$; as $L$ elicits the mean $\Gamma(p) = \E_p Y$, we have $\Var_p[Y] = \E_p (\E_pY - Y)^2 = \min_{r\in\reals} \E_p L(r,Y)$.
To apply Theorem~\ref{thm:bayes-risk-lower-bound}, we take this $L$ and $\Gamma$ and $d=1$, and must choose an appropriate level set $\Gamma_r$; we choose $p = \tfrac 1 n \ones$ to be the uniform distribution and let $r=\E_pY$.

Let us verify the three conditions:
(i) $p\in\Gamma_r$ by construction.
(ii) Letting $v\in\reals^\Y$ with $v_y = y - r$, define $F = \ker W \cap \affhull(\simplex)$ for $W = [v]\in\reals^{1\times n}$.
Note that $\Gamma_r \subseteq F$, and \raf{$W$ does not contain the $\ones$ direction}, as the $y$ values are distinct by assumption.
As $p\in F$, Lemma~\ref{lem:affhull-relint} gives $\affhull(\Gamma_r) = F$ and thus $\codim(\affhull(\Gamma_r)) = \rank(W) = 1 = d$.
(iii)
For $n \leq 2 = d+1$, $\Gamma_r=\{p\}$ is a singleton and we are done; otherwise $n\geq 3$.
\raft{I think we should punt this next bit to the appendix; it's pretty straightforward once you see what's happening, but really detracts from the simplicity of this example}
If $\Var[Y]$ were constant in $\Gamma_r$, then we would have some $c\in\reals$ such that $c = \Var_{p'}[Y] = \E_{p'} [Y^2] - r^2$ for all $p'\in\Gamma_r$.
Letting $W' = [v;v']\in\reals^{2\times n}$ where $v'_y = y^2-r^2-c$, and $F' = \ker W' \cap \affhull(\simplex)$, this would imply $\Gamma_r = F'\cap\simplex$ as well.
Lemma~\ref{lem:affhull-relint} applies again to show $F' = \affhull(\Gamma_r) = F$.
Yet as the $y$ values are distinct, $\rank(W')=2$ (for any $\alpha\in\reals$ there are at most two solutions to $y-r = \alpha(y^2-r^2-c)$), contradicting $\codim(F) = 1$.
Thus $\Var[Y]$ cannot be constant on $\Gamma_r$.

@@

For another application, let us see how Theorem~\ref{thm:bayes-risk-lower-bound} implies $\eliccvx(G) = n-1$ for any strictly concave function $G:\simplex\to\reals$ of the distribution, including most entropy functions.
In other words, there is no convex loss function allowing one to consistently estimate conditional entropy using fewer dimensions than required to estimate the conditional distribution itself.
This observation also extends to norms; given any $k>1$, $G(p) = \|p\|_k^k$ is strictly convex, and hence $\eliccvx(\|\cdot\|_k) = n-1$, as otherwise we could link to $G$ via $\psi(r) = r^k$.
These results illustrate the power of our technique.

@@

>>>>

%commented out Jun 1 2020
%Theorem~\ref{thm:consistent-implies-indir-elic} allows us to use convex elicitation complexity as a tool to understand efficiency of consistent convex surrogates for a given property, which is more often what is given in a continuous estimation setting.
%For example, when one wants to learn an $\alpha$-quantile, we start with the property rather than a loss.
%In the literature, pinball loss $L(r,y) = (r-y)(\ones_{r \geq y} - \alpha)$ typically appears without explanation or justification as to why it is relevant in relation to the $\alpha$-quantile.
%Elicitation teaches us why pinball loss is consistent for learning a quantile as the pinball loss elicits the $\alpha$-quantile.
In continuous estimation problems, often one does not start with a target loss per se, but instead with a conditional statistic of the data one wishes to estimate, such as the conditional mean, variance, or entropy.
In this setting, Theorem~\ref{thm:cvx-flats} gives lower bounds on the prediction dimension of convex losses with a link to the desired conditional statistic, i.e., the convex elicitation complexity.
In particular, following an argument from~\citet{frongillo2018elicitation}, below Theorem~\ref{thm:bayes-risk-lower-bound} yields new bounds on the convex elicitation complexity of properties quantifying risk or uncertainty such as variance and entropy.
These bounds address an open question posed by~\citet{frongillo2018elicitation}, that of developing a theory of elicitation complexity for convex losses; our results show that not only is this theory possible, but the bounds thus far remain the same for convex losses.
Extending to infinite $\Y$, such as $\Y=\reals$, remains open.
\raft{FYI to Jessie: what you had was all reasonable/true, but was a dispersed.  Which is understandable---just explaining the motivation for the edits.}

% Theorem~\ref{thm:cvx-flats} also addresses one major open question  from~\citet{frongillo2015elicitation,frongillo2018elicitation}\jessiet{which citation?}: \emph{what are lower bounds on convex elicitation complexity?}
% While Theorem~\ref{thm:cvx-flats} is a statement about consistent surrogates, the heart of it is actually about indirect property elicitation.
% Thus, the result also applies to extend previous results~\citep{frongillo2015elicitation,frongillo2018elicitation} to provide new bounds on elicitation complexity for certain classes of properties in Theorem~\ref{thm:bayes-risk-lower-bound}.

@@

  By Theorem~\ref{thm:cvx-flats}, as $\hat L$ elicits $\hat\Gamma$ and indirectly elicits $\Gamma$, there is a level set $\hat\Gamma_{\hat r}$ and flat $\hat F\subseteq\affhull(\simplex)$ containing $p$ such that $\codim(\hat F) \leq k$ and $S := \hat F \cap \simplex \subseteq \hat\Gamma_{\hat r} \subseteq \Gamma_r$.
  Let $F = \affhull(\Gamma_r)$ and recall $\codim(F)=d$.
  % Let $F' = \hat F \cap \affhull(\simplex)$, and observe the following: $p\in F'\cap\relint\simplex$, $F' \subseteq \affhull(\simplex)$, $S=F'\cap\simplex\subseteq\Gamma_r$, and $\codim(F') \leq k+1$.
  % the next \hat Fs used to be F's
  As $p\in \hat F\cap\relint\simplex$, Lemma~\ref{lem:affhull-relint} gives $\affhull(S) = \affhull(\hat F\cap \simplex) = \hat F$.
  Now as $S \subseteq \Gamma_r$, we also have $\hat F = \affhull(S) \subseteq \affhull(\Gamma_r) = F$,
  implying $\codim(\hat F) \geq \codim(F)$.

@@

\paragraph{Example: Variance.}
As a warm-up, let us see how to show $\eliccvx(\Var)=2$ whenever $|\Y|\geq 3$, meaning the lowest dimension of a convex loss to estimate conditional variance is 2.
It is interesting to note that, to the best of our knowledge, even this simple bound is novel.
While intuitively obvious, the lower bound of 2 is not trivial.
In particular, the well-known fact that the variance is not elicitable does not yield this lower bound, as it does not rule out the variance being a link of a real-valued convex-elicitable property; cf.~\citet{frongillo2018elicitation}.

Let $\Y\subseteq\reals$ be a finite set with $n=|\Y|$.
Observe that the variance of $Y$ is the Bayes risk of squared loss $L(r,y) = (r-y)^2$; as $L$ elicits the mean $\Gamma(p) = \E_p Y$, we have $\Var_p[Y] = \E_p (\E_pY - Y)^2 = \min_{r\in\reals} \E_p L(r,Y)$.
To apply Theorem~\ref{thm:bayes-risk-lower-bound}, we take this $L$ and $\Gamma$ and $d=1$, and must choose an appropriate level set $\Gamma_r$; we choose $p = \tfrac 1 n \ones$ to be the uniform distribution and let $r=\E_pY$.
We summarize three conditions here, leaving the full details to Appendix~\ref{app:variance-example}:
(i) $p\in\Gamma_r$ by construction.
(ii) Letting $v\in\reals^\Y$ with $v_y = y - r$, the flat $F = \ker W \cap \affhull(\simplex)$ for $W = [v]\in\reals^{1\times n}$ contains $\Gamma_r$, and has $\codim(\affhull(\Gamma_r)) = \rank(W) = 1 = d$.
(iii)
$\Gamma_r=\{p\}$ is a singleton for $n \leq 2 = d+1$, and for $n\geq 3$ there are enough degrees of freedom in $\simplex$ for there to be two distributions with mean $r$ and different variances.

@@

\paragraph{Example: Entropy and Norms.}
For another application, let us see how Theorem~\ref{thm:bayes-risk-lower-bound} implies $\eliccvx(G) = n-1$ for any strictly concave function $G:\simplex\to\reals$ of the distribution, including most entropy functions.
In other words, there is no convex loss function allowing one to consistently estimate conditional entropy using fewer dimensions than required to estimate the conditional distribution itself.
This observation also extends to norms; given any $k>1$, $G(p) = \|p\|_k^k$ is strictly convex, and hence $\eliccvx(\|\cdot\|_k) = n-1$, as otherwise we could link to $G$ via $\psi(r) = r^k$.
These results illustrate the power of our technique.

@@

\subsection{Variance example}
\label{app:variance-example}
We justify the three conditions of Theorem~\ref{thm:bayes-risk-lower-bound}:
\begin{enumerate}
\item[(i)]
  We have $p\in\Gamma_r$ by construction.
\item[(ii)]
  Let $\{y_1,\ldots,y_n\} = \Y$ be the $n$ distinct outcome/label values.
  Letting $v\in\reals^n$ with $v_i = y_i - r$, define $F = \ker W \cap \affhull(\simplex)$ for $W = [v]\in\reals^{1\times n}$.
  Note that $\Gamma_r \subseteq F$, and $\rank(W) = 1$ as the $y_i$ are distinct.
  As $p\in F$, Lemma~\ref{lem:affhull-relint} gives $\affhull(\Gamma_r) = F$ and thus $\codim(\affhull(\Gamma_r)) = \rank(W) = 1 = d$.
\item[(iii)]
  For $n \leq 2 = d+1$, $\Gamma_r=\{p\}$ is a singleton and we are done; otherwise $n\geq 3$.
  If $\Var[Y]$ were constant in $\Gamma_r$, then we would have some $c\in\reals$ such that $c = \Var_{p'}[Y] = \E_{p'} [Y^2] - r^2$ for all $p'\in\Gamma_r$.
  Letting $W' = [v;v']\in\reals^{2\times n}$ where $v'_y = y^2-r^2-c$, and $F' = \ker W' \cap \affhull(\simplex)$, this would imply $\Gamma_r = F'\cap\simplex$ as well.
  Lemma~\ref{lem:affhull-relint} applies again to show $F' = \affhull(\Gamma_r) = F$.
  Yet as the $y$ values are distinct, $\rank(W')=2$ (for any $\alpha\in\reals$ there are at most two solutions to $y-r = \alpha(y^2-r^2-c)$), contradicting $\codim(F) = 1$.
  Thus $\Var[Y]$ cannot be constant on $\Gamma_r$.
\end{enumerate}
